---
output:
  rmarkdown::github_document

params:
  redo_all:  TRUE
  redo_sync_ssu1:  TRUE
  redo_yaps_ssu1:  TRUE
---

<!-- README_sync.md is generated from README_sync.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README_sync-"
)
```
```{r, include=FALSE}
	
	redo_all 		<- params$redo_all
	if(redo_all){
		redo_sync_ssu1 <- redo_yaps_ssu1 <- TRUE
	} else {
		redo_sync_ssu1  <- params$redo_sync_ssu1
		redo_yaps_ssu1  <- params$redo_yaps_ssu1
	}
	
	if(!redo_sync_ssu1) {load(file="../readme_temps/sync_model_ssu1.rObj")}
	
	if(!redo_yaps_ssu1){
		load(file="../readme_temps/hydros_yaps_ssu1.rObj")
		load(file="../readme_temps/synced_dat_ssu1.rObj")
		load(file="../readme_temps/toa_ssu1.rObj")
		load(file="../readme_temps/inp_ssu1.rObj")
		load(file="../readme_temps/yaps_out_ssu1.rObj")
	}
	
```
# YAPS - How to use
### A complete walk-through taking you from raw data, through synchronization to final tracks
YAPS (Yet Another Position Solver) was introduced in 2017 as an alternative to proprietary and vendor specific software for estimating position and tracks of aquatic animals tagged with acoustic transmitters.  

Some motivations for developing and using YAPS:

* **Transparency and reproducibility.** These concepts should be fundamental in all science and scientific methodology. However, available proprietary software for estimating positions lack transparency and literally constitute “Black Boxes” making reproducibility very challenging, if not impossible.
	
* **Open source.** In addition to complete transparency, publishing YAPS open-source enables the user community to e.g. implement study specific adaptations, tweak parts of the estimation models and contribute to the continued development. Oh, and it is available free of charge. We urge users to submit improvements, refinements etc. to the main YAPS repository on github, thereby helping future users.
	
* **Vendor agnostic.** The basic input to YAPS is plain matrices of time-of-arrivals of transmitter signals detected by hydrophones. The brand of hydrophones and transmitters are irrelevant. This enables more direct comparisons of tracks and derived behavioural quantifications between studies using varying vendors.
	
* **Get the most out of your data.** To our knowledge, all vendor supplied tracking software process data from individual pings independently from other pings. By doing this, these algorithms explicitly discard collected data when pings are detected by less than three hydrophones. However, these data are valuable for estimating tracks if used correctly, and can dramatically increase overall data yield from your study (see e.g. Fig.6 in YAPS paper).
	
* **Get the best out of your data.** The holistic complete-track-oriented approach used by YAPS, have proven to be superior in many instances to the isolated single-ping-oriented approach used in vendor software. The single-ping approach is prone to suffer from numerical challenges, which can lead to varying degrees of spurious outliers, multiple position estimates from individual pings (sometimes known as “ghost positions”) and overall increased track jaggedness and uncertainties. These artefacts often necessitate the use of post-processing filters and/or smoothing techniques. In contrast, many of these irregularities are in most circumstances handled well by YAPS and resulting tracks are often ready for use in further analyses. Note, we are not stating that YAPS estimated tracks never need or will benefit from post-processing – it depends very much on study specifics such as the acoustic environment and transmitter burst interval type and duration. However, using YAPS estimated tracks as the starting point for post-processing have in all use cases we know of been vastly superior to using tracks estimated by vendor software.

Although YAPS development was focused on openness, transparency and flexibility, YAPS is not (nor has it been intended to be) a turn-key and/or a one-size-fits-all solution. While YAPS itself is open and transparent, the process from collecting raw data in the field to having final estimated tracks is long and sometimes challenging. Especially, the very important process of synchronizing arrays have proven to be a major obstacle for many potential users.  

This how-to guide is intended to provide users with an example work-flow to start from. The guide walks through the entire workflow from raw data collected by PPM-based systems (e.g. Vemco VR2 and Thelma TBR700) to final tracks. It shows how to use new functions added to YAPS enabling an easy and approachable (but fine tuned) synchronization. The guide does not cover any field specific steps in the work-flow (e.g. hydrophone array layout and fish tagging) as this information is available in other sources. We do advice users to make informed decisions regarding hydrophone array layout, to obtain precise positions of all hydrophones using high quality gps units and to reduce the potential for receiver movement during the study as much as possible.  


## Disclaimer
**YAPS obeys the fundamental rule of “garbage in, garbage out”. Therefore, DO NOT expect YAPS to salvage a poorly designed study, nor to turn crappy data into gold.**  
We have attempted to make both synchronization process and track estimation user-friendly. However, it is not trivial to synchronize hydrophones (let alone automating the process) based on detections in a variable and often noisy environment. Hydrophones might be replaced/shifted and if not fixed securely, hydrophones might move/be moved during a study. Additionally, hydrophone performance and output format varies considerably between (and within) manufacturers. On top of that, hydrophones don't always behave and perform as expected. For instance, some hydrophone models autonomously initiate reboots causing perturbation of varying magnitude and/or duration of the internal clock at apparently random time intervals. Therefore, the functions in YAPS might perform sub-optimal or even fail miserably when applied to new data. If/when this happens, please let us know through a direct message or leave a bug-report. Also note, the to-do list for improvements and tweaks is long and growing, so stay tuned for updates.  


### About synchronization in YAPS
Regardles of which positioning algorithm is used, synchronization of the hydrophone array prior to track estimation is *extremely* important for quality of the final tracks and great care should be taken to achieve as good as possible sync. The sync method included in YAPS has several benefits. First of all, it *synchronizes all hydrophones in one go*, avoiding sequential propagating synchronization that might lead to error accumulation in the outer ends of the array. Additionally, it allows for *estimation of hydrophone positions*, which can be usefull if the initial hydrophone positions are uncertain e.g. due to deep water, strong current or low-accuracy GPS. The internal clock in all hydrophones drift in different directions and the rate of drift is affected by water temperature. Therefore, temporally flexible non-linear correction is needed. The sync method employed by YAPS is based on a number of hydrophone-specific second-order polynomials. This whole-array-at-once, have proved to give very good synchronization. Perhaps most importantly, synchronization in YAPS is aimed at being user-friendly. To meet this, we have attempted to keep all the nitty-gritty stuff inside easy-to-use functions and kept the amount of user decisions at a minimum.

## Installation
Make sure you have the newest version of YAPS installed. For this, you need `devtools` installed - if not already installed, run `install.packages('devtools')`.  
YAPS relies heavily on use of Template Model Builder [TMB](https://github.com/kaskr/adcomp) for fitting the models, so make sure `TMB` is installed and working by following the simple [TMB instructions](https://github.com/kaskr/adcomp/wiki/Download).  
Then install the latest version of YAPS with:
```
	# install.packages("devtools")
	# install.packages("TMB")
	# TMB::runExample(all=TRUE)
	devtools::install_github("baktoft/yaps")
	library(yaps)
```

## Included example data
Two data sets are included in the package to serve as examples for how to use YAPS:  

1. **ssu1**: A tiny data set containing short test tracks from Florida Bay. 19 hydrophones, ~24 hour data, ~1 hour test track, Vemco VR2 PPM. This data set is very small and simple, 2D only and perfect for testing purposes. See`?ssu1`for further info.

2. **hald**: A larger data set from the Danish lake Hald Sø containing test and fish tracks. 70 hydrophones, 1 week data, ~1 hour test track, N fish tracks, Thelma TBR700 PPM. (Will be added).


### Data structure
Each data set is a list containing data.tables `hydros`, `detections` and `gps`. See e.g. `?ssu1` for further infomation.  
For help to create these tables from raw data, look in `?prepDetections` or [go here](#prepDetectionsHelp)
```{r example_hydros}
	head(ssu1$hydros, n=3)
	head(ssu1$detections, n=3)
	head(ssu1$gps, n=3)
```

## Workflow from raw to synchronized data

1. **Prepare data** list containing tables `hydros` and `detections` as described [here](#prepDetectionsHelp).

2. **Define sync parameters**. Each data set is different in terms of e.g. array configuration, precision and accuracy of hydrophone positioning, acoustic environment and detection probability, sync tag configuration, manufacturer etc. A number of parameters are available to setup the synchronization process for best results.
    + `max_epo_diff` Hydrophones are assumed to be running on at least somewhat similar time, so that ping trains from sync tags can be aligned correctly across all hydrophones. `max_epo_diff` sets the upper threshold for differences in TOA (Time Of Arrival) of a sync tag transmission detected at each hydrophone. Best parameter value depends on burst rate of sync tags and how far apart the internal clocks of the hydrophones are prior to synchronization. A bit less than half of minimum sync tag burst rate is a good starting choice. A linear time correction applied to the detection data prior to synchronization can improve the alignment process dramitically.
    
	+ `min_hydros` To ensure connectivity throughout the array, pings from sync tags need to be detected by multiple hydrophones and all hydrophones need to detect sync tags also detected by other hydrophones. `min_hydros` sets the lower threshold of how many hydrophones need to detect each sync tag ping in order to be included in the sync process. Should be as high as possible while observing that all hydrosphones are contributing. If too low, isolated hydrophones risk falling out completely. Future versions will work towards automation of this step.
    
	+ `time_keeper_idx` Index of the hydrophone to use as time keeper. Could e.g. be the one with smallest overall clock-drift.
	
	+ `fixed_hydros_idx` Vector of hydro idx's for all hydrophones where the position is assumed to be known with adequate accuracy and precision. Include as many as possible as fixed hydros to reduce overall computation time and reduce overall variability. As a bare minimum two hydros need to be fixed, but we strongly advice to use more than two.
	
	+ `n_offset_day` Specifies the number of hydrophone specific quadratic polynomials to use per day. For PPM based systems, 1 or 2 is often adeqaute.
	
	+ `n_ss_day` Specifies number of speed of sound to estimate per day. Future versions will enable use of logged water temperature instead. However, estimating SS gives an extra option for sanity-checking the final sync-model.
	
	+ `keep_rate` Syncing large data sets can take a long time. However, there is typically an excess number of sync tag detections and a sub-sample is typically enough for good synchronization. This parameter specifies the proportion (0-1) of data to keep when sub-sampling.  

3. **Compile input data** for the sync process using `getInpSync()`.

4. **Run** the sync model using `getSyncModel()`. This can take a long time for larger data sets including more hydrophones and/or covering longer time periods. It might be a good idea to start with the first few days to make sure everything looks ok before syncing the entire data set. Also consider the parameter `keep_rate` in `getInpSync()`.

5. **Check** the sync model to ensure the array is synced well using functions `plotSyncModelResids()` and `plotSyncModelCheck()`. These plots can be used to diagnose if the overall model fit is good.   
   Function `plotSyncModelResids()` plots sync model residuals (in meter) - sync model is ok if these residuals are centered closely around 0. If fixed hydrophones and/or sync tags consistently have large deviations from zero, it may indicate that position accuracy of the hydrophone is sub-optimal. Either obtain a better position (typically not possible) or allow th eposition to be non-fixed (i.e. remove from `fixed_hydros_idx`).  
   Function `plotSyncModelCheck()` apply the sync model to the sync data and compare true distances between hydros and sync tags to distances estimated based on the newly synced TOA-data. 

6. **Apply** the sync model to all data using `applySync()`.

The data should now be ready to be processed by YAPS to obtain estimated tracks.



## Processing example data ssu1
### Synchronizing - ssu1

1. Already taken care as we are using example data.

2. set the parameters...
```{r sync_ssu1_step2, warning=FALSE, results='hide', fig.show='hold', eval=redo_sync_ssu1}
	max_epo_diff <- 120
	min_hydros <- 2
	time_keeper_idx <- 5
	fixed_hydros_idx <- c(2:3,6, 8,11,13:17)
	n_offset_day <- 2
	n_ss_day <- 2
``` 

3. get input data ready for getSyncModel()...
```{r sync_ssu1_step3, warning=FALSE, results='hide', fig.show='hold', eval=redo_sync_ssu1}
	inp_sync <- getInpSync(sync_dat=ssu1, max_epo_diff, min_hydros, time_keeper_idx, fixed_hydros_idx, 
		n_offset_day, n_ss_day)
```

4. fit the sync model...
```{r sync_ssu1_step4, warning=FALSE, results='hide', fig.show='hold', eval=redo_sync_ssu1}
	sync_model <- getSyncModel(inp_sync, silent=TRUE)
```
```{r ssu1_save_sync_model, include=FALSE, eval=redo_sync_ssu1}
	save(sync_model, file="../readme_temps/sync_model_ssu1.rObj")
```


5. a   Plot model residuals (see code block below for explanation)...
```{r plotSyncModelResids, fig.show='hold', message=TRUE, fig.height=4, fig.width=4}
# Overall histogram of sync model residuals (converted to meter). 
#   Vertical red lines indicate 1 %, 5 %, 95 % and 99 % quantiles.
plotSyncModelResids(sync_model, by='overall') 

# Boxplots of sync model residuals (in meter) grouped by sync tags (panel) and hydrophone (x-axis). 
#  Vertical red lines added for every fifth hydro_idx to aid in identifying troublesome hydrophones. 
plotSyncModelResids(sync_model, by='sync_tag')

# Boxplots of sync model residuals (in meter) grouped by sync tags (x-axis) and hydrophone (panels).  
plotSyncModelResids(sync_model, by='hydro')
```

5. b   ...and model check plots. These plots apply `sync_model` to the data and compare *true* hydrophones <-> sync tags distances to *estimated* distances based on the synced data (delta). Closer to zero is better. See code block below for further info about each plot. 
```{r plotSyncModelCheck, fig.show='hold', message=FALSE, fig.height=4, fig.width=4}
# Delta grouped by sync tag (panel) and sync period (x-axis)
plotSyncModelCheck(sync_model, by="sync_bin_sync")

# Delta grouped by hydro_idx (panel) and sync period (x-axis)
plotSyncModelCheck(sync_model, by="sync_bin_hydro")

# Delta grouped by sync tag (panel) and hydro_idx (x-axis)
plotSyncModelCheck(sync_model, by="sync_tag")

# Delta grouped by hydro_idx (panel) and sync tag (x-axis)
plotSyncModelCheck(sync_model, by="hydro")
```

6. Apply the sync model to all detection data. Synced timestamps are found in column eposync
```{r applyingSync}
detections_synced <- applySync(toa=ssu1$detections, hydros=ssu1$hydros, sync_model)
head(detections_synced, n=3)
```

### Running YAPS - ssu1
Finally, run the synchronized data through YAPS - needs a little compiling beforehand.  
The only tracks to estimate from this data set are test tracks performed to test feasibility of deploying an array at this location. In the original data set, several transmitters with different power and burst rates were used. Here we focus on a single of these: ID 15266, Vemco V9, high power output, 20-40 second burst interval.
```{r runningYAPS, warning=FALSE, eval=redo_yaps_ssu1}
# We use the estimated positions of the hydros as they are assumed to be better
hydros_yaps <- data.table::data.table(sync_model$pl$TRUE_H)
colnames(hydros_yaps) <- c('hx','hy','hz')

# Specify focal tag and tag specific min and max burst intervals
focal_tag <- 15266
rbi_min <- 20
rbi_max <- 40

# Extract relevant data from the synced data
synced_dat_ssu1 <- detections_synced[tag == focal_tag]

# Compile TOA-matrix to use for YAPS
toa_ssu1 <- getToaYaps(synced_dat_ssu1, hydros_yaps, rbi_min, rbi_max)

# Compile all input data needed for YAPS
inp_ssu1 <- getInp(hydros_yaps, toa_ssu1, E_dist="Mixture", n_ss=2, pingType="rbi", sdInits=1, 
		rbi_min=rbi_min-5, rbi_max=rbi_max+5, ss_data_what="est", ss_data=0)

# Run YAPS to obtain estimated track
yaps_out_ssu1 <- runYaps(inp_ssu1, silent=TRUE) # Default parameter values should work for this example
```
```{r ssu1_save_yaps, include=FALSE, eval=redo_yaps_ssu1}
	save(hydros_yaps, file="../readme_temps/hydros_yaps_ssu1.rObj")
	save(synced_dat_ssu1, file="../readme_temps/synced_dat_ssu1.rObj")
	save(toa_ssu1, file="../readme_temps/toa_ssu1.rObj")
	save(inp_ssu1, file="../readme_temps/inp_ssu1.rObj")
	save(yaps_out_ssu1, file="../readme_temps/yaps_out_ssu1.rObj")
```



### Plot YAPS results - ssu1
Plot below shows true track (gps) as black hatched line while red line is YAPS tracking results. Green symbols indicate hydrophone positions.  

```{r plotYapsTrack, echo=FALSE, fig.height=6}
# Track estimate in pl
pl <- yaps_out_ssu1$pl
pl$X <- pl$X + inp_ssu1$inp_params$Hx0
pl$Y <- pl$Y + inp_ssu1$inp_params$Hy0
pl$top <- pl$top + inp_ssu1$inp_params$T0

# Error estimates in plsd
plsd <- yaps_out_ssu1$plsd

# plot the estimated track
plot(hy~hx, data=hydros_yaps, col="green", pch=20, cex=3, asp=1, xlab="UTM_X", ylab="UTM_Y")
lines(utm_y~utm_x, data=ssu1$gps, lty=2)
lines(pl$Y~pl$X, col="red")
```

X and Y coordinates of true track (black) and estimated track(red). Broken red lines indicate estimated track +- standard error of position estimate.  
```{r plotYapsOutDetails, echo=FALSE, fig.height=7}
par(mfrow=c(2,1))
plot(pl$top, pl$X, type="l", col="red", ylab="UTM_X", xlab="TimeOfPing", ylim=range(ssu1$gps$utm_x))
lines(utm_x~ts, data=ssu1$gps, lty=2)
lines(pl$top, pl$X-plsd$X, col="red", lty=3)
lines(pl$top, pl$X+plsd$X, col="red", lty=3)

plot(pl$top, pl$Y, type="l", col="red", ylab="UTM_Y", xlab="TimeOfPing", ylim=range(ssu1$gps$utm_y))
lines(utm_y~ts, data=ssu1$gps, lty=2)
lines(pl$top, pl$Y-plsd$Y, col="red", lty=3)
lines(pl$top, pl$Y+plsd$Y, col="red", lty=3)

```



***


### Caveats, notes and hints
+ Current version assumes all sync tags are co-located with hydrophones (which is usually the case). If needed, future versions will allow for other options.
+ For hydrophones where improved position is desired/needed (i.e. not in the `fixed_hydros_idx` vector), z is assumed to be known with much higher precision than x and y. 
+ Only use data from periods where entire array is up, stationary and running - otherwise, make sure to NA invalid data...
+ Ensuring initial accuracy and precision of hydrophone positions is as good as possible is probably the easiest way to improve final results. We recommend to use a differential GPS/GNSS unit instead of low-accuracy consumer grade devices.




***



## Appendix
#### How to create tables `hydros` and `detections` {#prepDetectionsHelp}
The `detections` tables contain all detections and can be constructed from csv-files exported from vendor software (millisecond or better time resolution is necessary; remember to enable this in export options, if needed). We advise to, if possible, apply a linear time correction on the detection data based on overall drift of hydrophone internal clock. Assuming the format follows the example shown below, the function `prepDetections()` can be used. Data should be truncated to only cover the period you want to synchronize - i.e. get rid of all detections before and after the study period. Also consider to exclude data recorded while the system is only partially in place - i.e. during installation, retrieval and/or downloading hydrophones. A minimum is to exclude data from each hydrophone in periods when it was not deployed at the recording position. Data removal is likely most easily done after import to R.   
There are many other formats out there - please tell us, if you want a specific format added to `prepDetections()`. Future versions will allow extraction of detections directly from the vendor provided SQLite database.
```{r example_detections}
	# Extract location of raw data included in yaps
	fn <- system.file("extdata", "VUE_Export_ssu1.csv", package="yaps") 
	vue <- data.table::fread(fn, fill=TRUE)
	head(vue, n=3)
	detections <- prepDetections(raw_dat=vue, type="vemco_vue")
	head(detections, n=3)
```

#### Example using YAPS on simulated data {#exampleSimuluation}
```
devtools::install_github("baktoft/yaps")
rm(list=ls())   
library(yaps)

# Simulate true track of animal movement of n seconds
trueTrack <- simTrueTrack(model='crw', n = 15000, deltaTime=1, shape=1, scale=0.5, addDielPattern=TRUE, ss='rw')

# Simulate telemetry observations from true track.
# Format and parameters depend on type of transmitter burst interval (BI) - stable (sbi) or random (rbi).
pingType <- 'sbi'

if(pingType == 'sbi') { # stable BI
    sbi_mean <- 30; sbi_sd <- 1e-4;
    teleTrack <- simTelemetryTrack(trueTrack, pingType=pingType, sbi_mean=sbi_mean, sbi_sd=sbi_sd)
} else if(pingType == 'rbi'){ # random BI
    pingType <- 'rbi'; rbi_min <- 20; rbi_max <- 40;
    teleTrack <- simTelemetryTrack(trueTrack, pingType=pingType, rbi_min=rbi_min, rbi_max=rbi_max)
}

# Simulate hydrophone array
hydros <- simHydros(auto=TRUE, trueTrack=trueTrack)
toa_list <- simToa(teleTrack, hydros, pingType, sigmaToa=1e-4, pNA=0.25, pMP=0.01)
toa <- toa_list$toa

# Specify whether to use ss_data from measured water temperature (ss_data_what <- 'data') or to estimate ss in the model (ss_data_what <- 'est')
ss_data_what <- 'data'
if(ss_data_what == 'data') {ss_data <- teleTrack$ss} else {ss_data <- 0}


if(pingType == 'sbi'){
    inp <- getInp(hydros, toa, E_dist="Mixture", n_ss=10, pingType=pingType, sdInits=0, ss_data_what=ss_data_what, ss_data=ss_data)
} else if(pingType == 'rbi'){
    inp <- getInp(hydros, toa, E_dist="Mixture", n_ss=10, pingType=pingType, sdInits=0, rbi_min=rbi_min, rbi_max=rbi_max, ss_data_what=ss_data_what, ss_data=ss_data)
} 
str(inp)

pl <- c()
maxIter <- ifelse(pingType=="sbi", 500, 5000)
outTmb <- runTmb(inp, maxIter=maxIter, getPlsd=TRUE, getRep=TRUE)
str(outTmb)

# Estimates in pl
pl <- outTmb$pl
# Correcting for hydrophone centering
pl$X <- outTmb$pl$X + inp$inp_params$Hx0
pl$Y <- outTmb$pl$Y + inp$inp_params$Hy0


# Error estimates in plsd
plsd <- outTmb$plsd

# plot the resulting estimated track
plot(y~x, data=trueTrack, type="l", xlim=range(hydros$hx), ylim=range(hydros$hy), asp=1)
lines(y~x, data=teleTrack)
points(hy~hx, data=hydros, col="green", pch=20, cex=3)
lines(pl$Y~pl$X, col="red")
```

