---
title: Tutorial 1 - installation, test and a quick example
output: rmarkdown::github_document
---
This tutorial is the first place to start venturing into `yaps`. In this, we will ensure everything is running as expected and introduce a general workflow and basic functionality of `yaps`. The tutorial was developed for the Ideas OTN workshop held in Halifax February 2020 and adapted slightly.

## Installation
`yaps` relies heavily on `TMB` (Template Model Builder). If not already installed, please install `TMB` and devtools. Then make sure you have the latest version of `yaps` installed. These steps are only needed once.

```{r, echo=TRUE, eval=FALSE}
install.packages('devtools')
install.packages("TMB", type = "source")
devtools::install_github('baktoft/yaps', ref='dev_ows')
```

Load libaries needed for this tutorial.
```{r, echo=TRUE, eval=TRUE, message=FALSE, warning=FALSE}
library(data.table)
library(dplyr)
library(sp)
library(leaflet)
```

Load `yaps` and test that everything is working as expected. This should display a short simulated known track (in black) and the track estimated by `yaps` in red.
```{r, echo=TRUE, eval=TRUE}
library(yaps)
testYaps(silent=TRUE)
```

## Getting started - the `ssu1` example data set

This is a tiny data set collected as part of a feasibility study using YAPS on Vemco PPM style data to track fish in shallow parts of Florida Bay, USA. Data collected by J.S. Rehage, J.R. Rodemann, R.S. Corujo and N. Viadero. Included in `yaps` with permission from [J.S. Rehage](https://case.fiu.edu/about/directory/profiles/rehage-jennifer-schopf.html), FIU Florida International University.

Have a look at the data - details can be found in `?ssu1`
```{r echo=TRUE}
names(ssu1)
head(ssu1$hydros)
```
Pretty self explanatory. Coordinates are in UTM - YAPS will (most probably) not work well with lat/lon data. Column `sync_tag` indicate serial number of special transmitters co-located with the hydrophones; data from these are used in the synchronization process. Column `idx` is an index running from `1:nrow(hydros)`.

```{r}
plot(y~x, data = ssu1$hydros, asp=1)
points(y~x, data=ssu1$hydros[!is.na(sync_tag)], col="red", pch=20)
```



```{r echo=TRUE}
head(ssu1$detections)
```
Almost self explanatory. Each row is a detection of a transmitter (`tag`) on a hydrophone identified by `serial`. Column `ts` is the (non-synced) timestamp of the detection in timezone UTC. Column `epo` is `ts` converted to UNIX epoch using `as.numeric(ts)` and `frac` is fractions of second for the detection, i.e. the complete time of detection is given by `epofrac = epo + frac`.

```{r echo=TRUE}
head(ssu1$gps)
```

Quick map to see where we are in the world (check out [`leaflet`](https://cran.r-project.org/web/packages/leaflet/index.html) - awesome for making quick slippy-maps)
```{r fig.height=25, echo=FALSE, message=FALSE}
hydros <- ssu1$hydros
coordinates(hydros) <- ~x+y
proj4string(hydros) <- CRS("+init=epsg:32617")
hydros_latlon <- spTransform(hydros, CRS("+init=epsg:4326"))
m <- leaflet(data=hydros_latlon, options = leafletOptions(minZoom = 0, maxZoom = 18), width="100%", height=700)
m <- addTiles(m, group="OSM")
m <- addCircles(m, radius=5, label=as.character(hydros_latlon$idx), labelOptions = labelOptions(noHide = T, textOnly = TRUE))
m <- addMeasure(m, primaryLengthUnit="meters")
m <- addProviderTiles(m, providers$Esri.WorldImagery, group="Esri.WorldImagery")
m <- addLayersControl(m, baseGroups = c("OSM (default)", "Esri.WorldImagery"),    options = layersControlOptions(collapsed = FALSE)  )
m
```


## Synchronizing the array
The code below is identical to that presented in our pre-print [Opening the black box of high resolution fish tracking using yaps](https://www.researchgate.net/publication/338010182_Opening_the_black_box_of_high_resolution_fish_tracking_using_yaps), which also include detailed description of the parameters in `getInpSync()`.

First, set the parameters to be used in the sync model and get input data prepared for use with `getSyncModel()`.
```{r eval=TRUE}
# set sync parameters 
max_epo_diff <- 120
min_hydros <- 2
time_keeper_idx <- 5
fixed_hydros_idx <- c(2:3, 6, 8, 11, 13:17)
n_offset_day <- 2
n_ss_day <- 2

inp_sync <- getInpSync(sync_dat=ssu1, max_epo_diff, min_hydros, time_keeper_idx, 
    fixed_hydros_idx, n_offset_day, n_ss_day, keep_rate=0.25)
```

Then, obtain a synchronization model using `getSyncModel()`
```{r eval=TRUE}
sync_model <- getSyncModel(inp_sync, silent=TRUE)
```

Use the diagnostic plots to ensure the obtained synchronization model is good. Basically, we want all number to be as close to zero as possible. Note, that a few outliers far away from zero is not serious, as the synchronization model follows a scaled t-distribution allowing long tails in the residuals.
```{r eval=TRUE}
plotSyncModelResids(sync_model, by='overall')    
plotSyncModelResids(sync_model, by='quantiles')
plotSyncModelResids(sync_model, by='sync_tag')      
plotSyncModelResids(sync_model, by='hydro')         
```


Next we apply the obtained synchronization model to the tracking data using `applySync()`
```{r eval=TRUE}
detections_synced <- applySync(toa=ssu1$detections, hydros=ssu1$hydros, sync_model)
```

## Running `yaps` to estimate the track
Now that we have the receivers synchronized, we can estimate the track using `yaps`. First, we need a data.table containing positions of the receivers in three dimension with colnum names 'hx', 'hy' and 'hz'.

```{r eval=TRUE}
hydros_yaps <- data.table::data.table(sync_model$pl$TRUE_H)
colnames(hydros_yaps) <- c('hx','hy','hz')
```

We specify tag specific minimum and maximum burst intervals and extract the detetections pertaining to the focal tag from the synchronized data
```{r eval=TRUE}
focal_tag <- 15266
rbi_min <- 20
rbi_max <- 40

synced_dat_ssu1 <- detections_synced[tag == focal_tag]
```

The main input data to `yaps` is a Time-of-Arrival matrix of the detections. This can be obtained using `getToaYaps()`
```{r eval=TRUE}
toa_ssu1 <- getToaYaps(synced_dat_ssu1, hydros_yaps, rbi_min, rbi_max)
```

Next, we use the function `getInp()` to compile input data for `runYaps()`.
```{r eval=TRUE}
inp_ssu1 <- getInp(hydros_yaps, toa_ssu1, E_dist="Mixture", n_ss=2, pingType="rbi", 
    sdInits=1, rbi_min=rbi_min, rbi_max=rbi_max, ss_data_what="est", ss_data=0)
```

Finally, we are ready to run `yaps` to obtain the track
```{r eval=TRUE}
yaps_out_ssu1 <- runYaps(inp_ssu1, silent=TRUE)
```

## Basic plotting of estimated track
To do some very basic visualization of the obtained track, we can use `plotYaps()` as below.
```{r eval=TRUE}
plotYaps(yaps_out=yaps_out_ssu1, type="map")
# Add gps track for direct comparison
lines(utm_y~utm_x, data=ssu1$gps, lty=2)

par(mfrow=c(2,1))
plotYaps(yaps_out=yaps_out_ssu1, type="coord_X")
lines(utm_x~ts, data=ssu1$gps, lty=2)
plotYaps(yaps_out=yaps_out_ssu1, type="coord_Y")
lines(utm_y~ts, data=ssu1$gps, lty=2)
```

This concludes the first tutorial on using `yaps`. 


